\titletext{}

### 2021-09-15 manuscript/regional_befr_manuscript.tex
Updated bioclim explan
Added published paper

Plots were chosen from the SEOSAW database based on the quality and completeness of data collection, and plot configuration. Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands, or savannas \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction, experimental treatments such as prescribed burning or herbivore exclusion, or containing termite mounds were excluded from the initial pool. Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate biomass estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered here a separate biome with very different species composition \citep{Parr2014}. \nzam{} plots within the SEOSAW database were arranged in clusters of four 20x50 m plots, with 20 m between plots. Plots within each spatial cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.

@@

Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}. In addition to MAT and MAP, temperature stress was calculated as the mean diurnal temperature range (BIO2) and precipitation seasonality was calculated as the mean of the coefficient of variation of monthly mean precipitation (BIO15). Soil fertility data were extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC) (cmolc kg\textsuperscript{-1}), soil organic carbon stocks (kg m\textsuperscript{-2}) percentage soil sand content (0.05-2 mm) by weight and soil nitrogen content (g kg\textsuperscript{-1}). These data are a modelled product derived from various remotely sensed and directly measured data sources. The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1, \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018. We initially aimed to include disturbance by herbivory in our model, including total herbivore biomass from the \citet{Hempson2017} modelled herbivory product, but this inclusion prevented models from converging due to its collinearity with other observed variables, notably MAP and disturbance by fire.

@@

Estimated tree species richness was calculated for each plot using `ChaoRichness()' from the `iNEXT' package in R \citep{Hsieh2016}. This procedure uses Hill numbers of the order $q=0$ to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value \citep{Jost2006}. Extrapolated species richness accounts for variation in plot size present in the dataset (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}. To measure tree species evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness. Abundance evenness allows for greater niche complementarity at small scales due to potentially increased heterogeneity of functional traits. In terms of the theory of niche complementarity, both species richness and abundance evenness contribute to the positive total diversity effect on ecosystem function. Holding either species richness or abundance evenness constant while increasing the other will have the similar effect of reducing the likelihood that a neighbour is a conspecific, thus reducing the occurrence of negative density dependent competition effects and therefore increasing ecosystem function. We quantified tree structural diversity for each plot by calculating the Coefficient of Variation of DBH (DBH CoV) and tree height (Height CoV).

@@

Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficients of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Figure S3). Stem density had the strongest correlation with AGB of all variables considered (\ccib{}). Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency. The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs. MAP had positive correlations with all tree species diversity and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).

>>>>

Plots were chosen from the SEOSAW database based on the quality and completeness of data collection, and plot configuration. Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands, or savannas \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction, experimental treatments such as prescribed burning or herbivore exclusion, or containing termite mounds were excluded from the initial pool. Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate biomass estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered here a separate biome with very different species composition \citep{Parr2014}. \nzam{} plots within the SEOSAW database were arranged in clusters of four 20$\times$50 m plots, with 20 m between plots. Plots within each spatial cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.

@@

Climatic data were taken from the WorldClim database \citep{Fick2017}, which uses BIOCLIM to derive bioclimatic variables \citep{Booth2014}. In addition to MAT and MAP, temperature stress was calculated as the mean diurnal temperature range (BIO2) and precipitation seasonality was calculated as the mean of the coefficient of variation of monthly mean precipitation (BIO15). Soil fertility data were extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC) (cmolc kg\textsuperscript{-1}), soil organic carbon stocks (kg m\textsuperscript{-2}) percentage soil sand content (0.05-2 mm) by weight and soil nitrogen content (g kg\textsuperscript{-1}). These data are a modelled product derived from various remotely sensed and directly measured data sources. The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1, \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018. We initially aimed to include disturbance by herbivory in our model, including total herbivore biomass from the \citet{Hempson2017} modelled herbivory product, but this inclusion prevented models from converging due to its collinearity with other observed variables, notably MAP and disturbance by fire.

@@

Estimated tree species richness was calculated for each plot using `ChaoRichness()' from the `iNEXT' package in R \citep{Hsieh2016}. This procedure uses Hill numbers of the order $q=0$ to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value \citep{Jost2006}. Extrapolated species richness accounts for variation in plot size present in the dataset (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}. To measure tree species evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness. Abundance evenness allows for greater niche complementarity at small scales due to potentially increased heterogeneity of functional traits. In terms of the theory of niche complementarity, both species richness and abundance evenness contribute to the positive total diversity effect on ecosystem function. Holding either species richness or abundance evenness constant while increasing the other will have the similar effect of reducing the likelihood that a neighbour is a conspecific, thus reducing the occurrence of negative density dependent competition effects and therefore increasing ecosystem function. We quantified tree structural diversity for each plot by calculating the Coefficient of Variation of DBH (DBH CV) and tree height (Height CV).

@@

Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficients of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Figure S3). Stem density had the strongest correlation with AGB of all variables considered (\ccib{}). Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency. The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs. MAP had positive correlations with all tree species diversity and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CV: \ccdvi{}, Height CV: \cchvi{}).
